---
title: "CGDA - Case Study 2 - Analyze Phase"
output: html_document
date: "2023-04-10"
---

```{r set outputs, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results = 'hide')
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

# Analyze Data

## Load Packages
First we'll load in packages for the processing phase.

- tidyverse: a collection of data analysis packages
- skimr: provides summary statistics
```{r load packages}
library(tidyverse)
library(skimr)
```

```{r update outputs, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results = 'markup')
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning = TRUE)
knitr::opts_chunk$set(message = TRUE)
```

## Load CSV Files
We'll load the CSV files into separate data frames for analysis. We'll also set any ordered factors.
```{r load CSV files into dataframes}
df_daily <- read_csv("daily.csv", col_types=cols(id=col_character(), date = col_date(format="%Y-%m-%d"), cs_id=col_character()))
df_daily$cs_id <- ordered(df_daily$cs_id, levels=c('1', '2', '3', '4', '5', '6', '7', '8', '9', '10', '11', '12', '13', '14', '15', '16', '17', '18', '19', '20', '21', '22', '23', '24', '25', '26', '27', '28', '29', '30', '31', '32', '33'))
df_hourly <- read_csv("hourly.csv", col_types=cols(id=col_character(), date_time = col_date(format="%Y-%m-%d %H:%M:%S"),date = col_date(format="%Y-%m-%d")))
df_minutes <- read_csv("minutes.csv", col_types=cols(id=col_character(), date_time = col_date(format="%Y-%m-%d %H:%M:%S")))
df_sleep_daily <- read_csv("sleep_daily.csv",col_types=cols(id=col_character(), date = col_date(format="%Y-%m-%d")))
df_sleep_minutes <- read_csv("sleep_minutes.csv", col_types=cols(id=col_character(), date_time = col_date(format="%Y-%m-%d %H:%M:%S")))
```

## Analyze Daily Data
We'll start by  analyzing the daily data. First we'll take a look at the data structure, then we'll plot a few items based on the available data.

### Inspect and Summarize Daily Data
```{r analyze daily d}
str(df_daily)
summary(df_daily)
```

### Plot Daily Data
While it is intriguing to analyze collected data for health or statistical correlation, we must remember that the goal is to identify trends in usage of the devices. It is easy to become distracted with creating many graphs that may not help solve the business task. We are going to focus on analyzing the usage of the devices to see if we can develop relevant insights.
```{r plot daily data}
# Total records by id
df_daily %>%
  group_by(cs_id) %>%
  summarize(total_records = n()) %>%
  ggplot(aes(x=cs_id, y=total_records, fill=cs_id)) + geom_col() + scale_fill_discrete()

# Total minutes of recording by id
# there are 43,200 minutes in 30 days
df_daily %>%
  group_by(cs_id) %>%
  summarize(total_minutes = sum(very_active_minutes, fairly_active_minutes, lightly_active_minutes, sedentary_minutes)) %>%
  ggplot(aes(x=cs_id, y=total_minutes, fill=cs_id)) + geom_col() + scale_fill_discrete()

# Total minutes recorded each day, line = id
# there are 1,440 minutes in 1 day
df_daily %>%
  rowwise() %>%
  mutate(daily_minutes = sum(very_active_minutes, fairly_active_minutes, lightly_active_minutes, sedentary_minutes)) %>%
  ggplot(aes(x=date, y=daily_minutes, color=cs_id)) + geom_line() + facet_wrap(~cs_id) + scale_color_discrete() + theme(axis.text.x = element_text(angle=90))

# Total minutes recorded each day, by weekday
df_daily %>%
  mutate(weekday = wday(date, label=TRUE)) %>%
  group_by(weekday) %>%
  summarise(total_minutes = sum(very_active_minutes, fairly_active_minutes, lightly_active_minutes, sedentary_minutes)) %>%
  ggplot(aes(x=weekday, y=total_minutes, fill=weekday)) + geom_col() + scale_fill_discrete()

# Total minutes recorded each weekday, by id
df_daily %>%
  mutate(weekday = wday(date, label=TRUE)) %>%
  group_by(cs_id, weekday) %>%
  summarise(total_minutes = sum(very_active_minutes, fairly_active_minutes, lightly_active_minutes, sedentary_minutes)) %>%
  ggplot(aes(x=weekday, y=total_minutes, fill=cs_id)) + geom_col() + facet_wrap(~cs_id) + scale_fill_discrete() + theme(axis.text.x = element_text(angle=90))

# Total steps recorded each weekday, by id
df_daily %>%
  mutate(weekday = wday(date, label=TRUE)) %>%
  group_by(cs_id, weekday) %>%
  summarise(sum_steps = sum(total_steps)) %>%
  ggplot(aes(x=weekday, y=sum_steps, fill=cs_id)) + geom_col() +
  facet_wrap(~cs_id) + scale_fill_discrete() + theme(axis.text.x = element_text(angle=90))
```