---
title: "CGDA - Case Study 2 - Analyze Phase"
output: html_notebook
date: "2023-04-10"
---

```{r set outputs, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results = 'hide')
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

# Analyze Data

## Load Packages
First we'll load in packages for the processing phase.

- tidyverse: a collection of data analysis packages
- skimr: provides summary statistics
```{r load packages}
library(tidyverse)
library(skimr)
```

```{r update outputs, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results = 'markup')
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning = TRUE)
knitr::opts_chunk$set(message = TRUE)
```

## Load CSV Files
We'll load the CSV files into separate data frames for analysis. We'll also set any ordered factors.
```{r load CSV files into dataframes}
df_daily <- read_csv("daily.csv", col_types=cols(id=col_character(), date = col_date(format="%Y-%m-%d"), cs_id=col_character()))
df_daily$cs_id <- ordered(df_daily$cs_id, levels=c('1', '2', '3', '4', '5', '6', '7', '8', '9', '10', '11', '12', '13', '14', '15', '16', '17', '18', '19', '20', '21', '22', '23', '24', '25', '26', '27', '28', '29', '30', '31', '32', '33'))
df_hourly <- read_csv("hourly.csv", col_types=cols(id=col_character(), date_time = col_datetime(format="%Y-%m-%d %H:%M:%S"),date = col_date(format="%Y-%m-%d")))
df_hourly$cs_id <- ordered(df_hourly$cs_id, levels=c('1', '2', '3', '4', '5', '6', '7', '8', '9', '10', '11', '12', '13', '14', '15', '16', '17', '18', '19', '20', '21', '22', '23', '24', '25', '26', '27', '28', '29', '30', '31', '32', '33'))
df_minutes <- read_csv("minutes.csv", col_types=cols(id=col_character(), date_time = col_datetime(format="%Y-%m-%d %H:%M:%S")))
df_minutes$cs_id <- ordered(df_minutes$cs_id, levels=c('1', '2', '3', '4', '5', '6', '7', '8', '9', '10', '11', '12', '13', '14', '15', '16', '17', '18', '19', '20', '21', '22', '23', '24', '25', '26', '27', '28', '29', '30', '31', '32', '33'))
df_sleep_daily <- read_csv("sleep_daily.csv",col_types=cols(id=col_character(), date = col_date(format="%Y-%m-%d")))
df_sleep_daily$cs_id <- ordered(df_sleep_daily$cs_id, levels=c('1', '2', '3', '4', '5', '6', '7', '8', '9', '10', '11', '12', '13', '14', '15', '16', '17', '18', '19', '20', '21', '22', '23', '24', '25', '26', '27', '28', '29', '30', '31', '32', '33'))
df_sleep_minutes <- read_csv("sleep_minutes.csv", col_types=cols(id=col_character(), date_time = col_datetime(format="%Y-%m-%d %H:%M:%S")))
```

## Analyze Daily Data
We'll start by  analyzing the daily data. First we'll take a look at the data structure, then we'll plot a few items based on the available data.

### Inspect and Summarize Daily Data
```{r analyze daily data}
str(df_daily)
summary(df_daily)
```

### Plot Daily Data
While it is intriguing to analyze collected data for health or statistical correlation, we must remember that the goal is to identify trends in usage of the devices. It is easy to become distracted with creating many graphs that may not help solve the business task. We are going to focus on analyzing the usage of the devices to see if we can develop relevant insights.
```{r plot daily data}
# Total records by id
df_daily %>%
  group_by(cs_id) %>%
  summarize(total_records = n()) %>%
  ggplot(aes(x=cs_id, y=total_records, fill=cs_id)) + geom_col() + scale_fill_discrete() + labs(title="Total Daily Records per Participant") + xlab("Participant ID") + ylab("Total Records")

# Total minutes of recording by id
# there are 43,200 minutes in 30 days
df_daily %>%
  group_by(cs_id) %>%
  summarize(total_minutes = sum(very_active_minutes, fairly_active_minutes, lightly_active_minutes, sedentary_minutes)) %>%
  ggplot(aes(x=cs_id, y=total_minutes, fill=cs_id)) + geom_col() + scale_fill_discrete() + labs(title="Total Minutes of Recording per Participant") + xlab("Participant ID") + ylab("Total Minutes")

# Total minutes recorded each day, line = id
# there are 1,440 minutes in 1 day
df_daily %>%
  rowwise() %>%
  mutate(daily_minutes = sum(very_active_minutes, fairly_active_minutes, lightly_active_minutes, sedentary_minutes)) %>%
  ggplot(aes(x=date, y=daily_minutes, color=cs_id)) + geom_line() + facet_wrap(~cs_id) + scale_color_discrete() + theme(axis.text.x = element_text(angle=90)) + labs(title="Total Minutes Recorded Each Day by Participant") + xlab("Date") + ylab("Total Daily Minutes")

# Total minutes recorded each day, by weekday
df_daily %>%
  mutate(weekday = wday(date, label=TRUE)) %>%
  group_by(weekday) %>%
  summarise(total_minutes = sum(very_active_minutes, fairly_active_minutes, lightly_active_minutes, sedentary_minutes)) %>%
  ggplot(aes(x=weekday, y=total_minutes, fill=weekday)) + geom_col() + scale_fill_discrete() + labs(title="Total Minutes Recorded each Weekday by All Participants") + xlab("Weekday") + ylab("Total Daily Minutes")

# Total minutes recorded each weekday, by id
df_daily %>%
  mutate(weekday = wday(date, label=TRUE)) %>%
  group_by(cs_id, weekday) %>%
  summarise(total_minutes = sum(very_active_minutes, fairly_active_minutes, lightly_active_minutes, sedentary_minutes)) %>%
  ggplot(aes(x=weekday, y=total_minutes, fill=cs_id)) + geom_col() + facet_wrap(~cs_id) + scale_fill_discrete() + theme(axis.text.x = element_text(angle=90))

# Average minutes recorded each weekday, by id
df_daily %>%
  mutate(weekday = wday(date, label=TRUE)) %>%
  mutate(total_minutes = very_active_minutes+fairly_active_minutes+lightly_active_minutes+sedentary_minutes) %>%
  group_by(cs_id, weekday) %>%
  summarise(total_minutes = mean(total_minutes)) %>%
  ggplot(aes(x=weekday, y=total_minutes, fill=cs_id)) + geom_col() + facet_wrap(~cs_id) + scale_fill_discrete() + theme(axis.text.x = element_text(angle=90))

# Total steps recorded each weekday, by id
df_daily %>%
  mutate(weekday = wday(date, label=TRUE)) %>%
  group_by(cs_id, weekday) %>%
  summarise(sum_steps = sum(total_steps)) %>%
  ggplot(aes(x=weekday, y=sum_steps, fill=cs_id)) + geom_col() +
  facet_wrap(~cs_id) + scale_fill_discrete() + theme(axis.text.x = element_text(angle=90))

# Average steps recorded each weekday, by id
df_daily %>%
  mutate(weekday = wday(date, label=TRUE)) %>%
  group_by(cs_id, weekday) %>%
  summarise(avg_steps = mean(total_steps)) %>%
  ggplot(aes(x=weekday, y=avg_steps, fill=cs_id)) + geom_col() + facet_wrap(~cs_id) + scale_fill_discrete()

# Scatter plot of steps to sedentary minutes
df_daily %>%
  ggplot(aes(x=total_steps, y=sedentary_minutes)) + geom_point() + scale_color_discrete() + geom_smooth()

# Scatter plot of steps to very active minutes
df_daily %>%
  ggplot(aes(x=total_steps, y=very_active_minutes)) + geom_point() + scale_color_discrete() + geom_smooth()

# Pie plot of activity levels for all ids
df_daily %>%
  summarise('Very Active'=sum(very_active_minutes), 'Fairly Active'=sum(fairly_active_minutes),'Lightly Active'=sum(lightly_active_minutes),'Sedentary'=sum(sedentary_minutes)) %>%
  pivot_longer(cols=names(.)) %>%
  ggplot(aes(x="", y=value, fill=name)) + geom_bar(stat='identity', width=1) + coord_polar("y", start=0) + theme_void() + scale_fill_discrete()

# average daily intensity, by id
df_daily %>%
  group_by(cs_id) %>%
  summarise('Very Active' = mean(very_active_minutes), 'Fairly Active' = mean(fairly_active_minutes), 'Lightly Active' = mean(lightly_active_minutes), 'Sedentary' = mean(sedentary_minutes)) %>%
  pivot_longer(cols='Very Active':'Sedentary') %>%
  ggplot(aes(x=cs_id, y=value, fill=name)) + geom_bar(stat='identity') + labs(title="Average Daily Intensity", x="Participant", y="Average Daily Intensity") + theme(axis.text.x = element_text(angle=90))

# average distance by weekday, by id
df_daily %>%
  mutate(weekday = wday(date, label=TRUE)) %>%
  group_by(cs_id, weekday) %>%
  summarise(avg_dist = mean(total_distance)) %>%
  ggplot(aes(x=weekday, y=avg_dist, fill=cs_id)) + geom_col() + facet_wrap(~cs_id) + scale_fill_discrete() + theme(axis.text.x = element_text(angle=90))

# total distance each day, by id
df_daily %>%
  ggplot(aes(x=date, y=total_distance, color=cs_id)) + geom_line() + facet_wrap(~cs_id) + scale_color_discrete() + theme(axis.text.x = element_text(angle=90)) + labs(title="Total Distance Recorded Each Day by Participant") + xlab("Date") + ylab("Total Daily Distance (km)")
```

```{r plot daily sleep data}
# scatter plot total time in bed vs total minutes asleep
df_sleep_daily %>%
  ggplot(aes(x=total_time_in_bed, y=total_minutes_asleep)) + geom_point() + geom_smooth()

# time in bed vs minutes asleep, by id
df_sleep_daily %>%
  ggplot(aes(x=total_time_in_bed, y=total_minutes_asleep, color=cs_id)) + geom_point() + facet_wrap(~cs_id) + scale_color_discrete()

# average time in bed by weekday, by id
df_sleep_daily %>%
  mutate(weekday = wday(date, label=TRUE)) %>%
  group_by(cs_id, weekday) %>%
  summarise(avg_bed = mean(total_time_in_bed)) %>%
  ggplot(aes(x=weekday, y=avg_bed, fill=cs_id)) + geom_col() + facet_wrap(~cs_id) + scale_fill_discrete()+ theme(axis.text.x = element_text(angle=90))

# average time asleep by weekday, by id
df_sleep_daily %>%
  mutate(weekday = wday(date, label=TRUE)) %>%
  group_by(cs_id, weekday) %>%
  summarise(avg_sleep = mean(total_minutes_asleep)) %>%
  ggplot(aes(x=weekday, y=avg_sleep, fill=cs_id)) + geom_col() + facet_wrap(~cs_id) + scale_fill_discrete()+ theme(axis.text.x = element_text(angle=90))
```

## Analyze Hourly Data

### Inspect and Summarize Hourly Data
```{r inspect hourly data}
str(df_hourly)
summary(df_hourly)
```

### Plot Hourly Data
Intensity values are summarizations of by-the-minute activity level records. For example, as will be seen in the df_minutes data frame, the smart device labeled each minute as Very Active/Fairly Active/Lightly Active/Sedentary from 3 to 0 based on proprietary algorithms. The total_intensity field sums all of the individual minutes' activity level (0-3). The average intensity field divides the total_intensity field by 60 (minutes) to determine the average activity level throughout the hour.
```{r plot hourly data}
# Total records by id
df_hourly %>%
  group_by(cs_id) %>%
  summarise(total_records = n()) %>%
  ggplot(aes(x=cs_id, y=total_records, fill=cs_id)) + geom_col() + scale_fill_discrete() + labs(title='Total Hourly Records per Participant', x='Participant ID', y='Total Records')

# average hourly total intensity across all ids
df_hourly %>%
  mutate(hour = hour(date_time)) %>%
  group_by(hour) %>%
  summarise(avg_intensity = mean(total_intensity)) %>%
  ggplot(aes(x=hour, y=avg_intensity)) + geom_col() + scale_fill_discrete()

# average hourly intensity across all ids
df_hourly %>%
  mutate(hour = hour(date_time)) %>%
  group_by(hour) %>%
  summarise(avg_intensity = mean(average_intensity)) %>%
  ggplot(aes(x=hour, y=avg_intensity)) + geom_col() + scale_fill_discrete()

# average hourly total intensity, by id
df_hourly %>%
  mutate(hour = hour(date_time)) %>%
  group_by(cs_id, hour) %>%
  summarise(avg_intensity = mean(total_intensity)) %>%
  ggplot(aes(x=hour, y=avg_intensity, fill=cs_id)) + geom_col() + facet_wrap(~cs_id) + scale_fill_discrete()

# total intensity value by the hour, by id
df_hourly %>%
  ggplot(aes(x=date_time, y=total_intensity, color=cs_id)) + geom_line() + facet_wrap(~cs_id) + scale_color_discrete() + theme(axis.text.x = element_text(angle=90)) + labs(title='Total Intensity by the Hour per Participant', x='Date/Time', y='Hourly Intensity')
```

## Analyze Minutes Data

### Inspect and Summarize Minutes Data
```{r inspect minutes data}
str(df_minutes)
summary(df_minutes)
```

### Plot Minutes Data
```{r plot minutes data}
# average recordings by the hour
df_minutes %>%
  mutate(hour = hour(date_time)) %>%
  group_by(cs_id, hour) %>%
  summarise(count_hour = n()) %>%
  ggplot(aes(x=hour, y=count_hour, color=cs_id)) + geom_line() + facet_wrap(~cs_id) + scale_color_discrete()

# total active (>0/not sedentary) intensity recorded by weekday, per id
df_minutes %>%
  filter(intensity > 0) %>%
  mutate(weekday = wday(date_time, label=TRUE)) %>%
  group_by(cs_id, weekday) %>%
  summarise(sum_intensity = sum(intensity)) %>%
  ggplot(aes(x=weekday, y=sum_intensity, fill=cs_id)) + geom_col() + facet_wrap(~cs_id) + scale_fill_discrete() + theme(axis.text.x = element_text(angle=90))

# total calories burned by weekday, per id
df_minutes %>%
  mutate(weekday = wday(date_time, label=TRUE)) %>%
  group_by(cs_id, weekday) %>%
  summarise(sum_mets = sum(calories)) %>%
  ggplot(aes(x=weekday, y=sum_mets, fill=cs_id)) + geom_col() + facet_wrap(~cs_id) + scale_fill_discrete() + theme(axis.text.x = element_text(angle=90))

# average intensity by the hour per all ids
df_minutes %>%
  mutate(hour = hour(date_time)) %>%
  group_by(hour) %>%
  summarise(avg_intensity = mean(intensity)) %>%
  ggplot(aes(x=hour, y=avg_intensity)) + geom_col()

# average intensity by the hour, by id
df_minutes %>%
  mutate(hour = hour(date_time)) %>%
  group_by(cs_id, hour) %>%
  summarise(avg_intensity = mean(intensity)) %>%
  ggplot(aes(x=hour, y=avg_intensity, fill=cs_id)) + geom_col() + facet_wrap(~cs_id) + scale_fill_discrete()

# average intensity by the weekday and hour
df_minutes %>%
  mutate(hour = hour(date_time), weekday = wday(date_time, label=TRUE)) %>%
  group_by(weekday, hour) %>%
  summarise(avg_intensity = mean(intensity)) %>%
  ggplot(aes(x=hour, y=avg_intensity)) + geom_col() + facet_wrap(~weekday)

# average intensity by the weekday and hour, per id
# we will look at just a few ids due to poor rendering if we do all at once
df_minutes %>%
  filter(cs_id == '1'|cs_id == '10'|cs_id == '20'|cs_id == '30') %>%
  mutate(hour = hour(date_time), weekday = wday(date_time, label=TRUE)) %>%
  group_by(cs_id, weekday, hour) %>%
  summarise(avg_intensity = mean(intensity)) %>%
  ggplot(aes(x=hour, y=avg_intensity, fill=cs_id)) + geom_col() + scale_fill_discrete() + facet_grid(cs_id ~ weekday)

# average steps by the weekday and hour, per id
# we will look at just a few ids due to poor rendering if we do all at once
df_minutes %>%
  filter(cs_id == '1'|cs_id == '10'|cs_id == '20'|cs_id == '30') %>%
  mutate(hour = hour(date_time), weekday = wday(date_time, label=TRUE)) %>%
  group_by(cs_id, weekday, hour) %>%
  summarise(avg_steps = mean(steps)) %>%
  ggplot(aes(x=hour, y=avg_steps, fill=cs_id)) + geom_col() + scale_fill_discrete() + facet_grid(cs_id ~ weekday)
```