---
title: "CGDA - Case Study 1 - Process Data"
output: html_document
date: "2023-03-28"
---
# Process the Data
We will begin the process phase by loading, transforming, and validating the data.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results = 'hide')
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

## Load Packages
```{r load packages}
library(tidyverse)
```

## Load CSV Files
```{r load CSV files}
d202203 <- read_csv("202203-divvy-tripdata.csv")
d202204 <- read_csv("202204-divvy-tripdata.csv")
d202205 <- read_csv("202205-divvy-tripdata.csv")
d202206 <- read_csv("202206-divvy-tripdata.csv")
d202207 <- read_csv("202207-divvy-tripdata.csv")
d202208 <- read_csv("202208-divvy-tripdata.csv")
d202209 <- read_csv("202209-divvy-publictripdata.csv")
d202210 <- read_csv("202210-divvy-tripdata.csv")
d202211 <- read_csv("202211-divvy-tripdata.csv")
d202212 <- read_csv("202212-divvy-tripdata.csv")
d202301 <- read_csv("202301-divvy-tripdata.csv")
d202302 <- read_csv("202302-divvy-tripdata.csv")
```

## Quick Look
Let's take a quick look at a dataframe:
```{r quick look, results='markup'}
colnames(d202203)
spec(d202203)
str(d202203)
head(d202203)
summary(d202203)
```

## Compare Column Names and Data Types
Before we can begin processing the data, let's verify the column names and data types match across the data frames.

```{r compare data frames, results='markup'}
spec(d202203)
spec(d202204)
spec(d202205)
spec(d202206)
spec(d202207)
spec(d202208)
spec(d202209)
spec(d202210)
spec(d202211)
spec(d202212)
spec(d202301)
spec(d202302)
```

## Combine Data Frames
Combine data into a single data frame for further processing.
```{r combine data frames, results='markup'}
all_rides <- bind_rows(d202203, d202204, d202205, d202206, d202207, d202208, d202209, d202210, d202211, d202212, d202301, d202302)
```

### Clean Up Memory
Remove the individual data frames from memory since we are done using them.
```{r clean memory}
rm(d202203, d202204, d202205, d202206, d202207, d202208, d202209, d202210, d202211, d202212, d202301, d202302)
gc()
```

## Inspect the New Data Frame
We'll verify the structure and sample data from the new data frame.
```{r inspect combined data frame, results='markup'}
colnames(all_rides)
spec(all_rides)
str(all_rides)
head(all_rides)
summary(all_rides)
```

## Remove Unused Data
We'll filter out unused columns of data and save it to a new data frame for further processing.
```{r remove unused data}
filtered_rides <- all_rides %>%
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

## Inspect the Data Frame Again
We conduct frequent verification of the changes that have been made.
```{r verify filtered data, results='markup'}
colnames(filtered_rides)
nrow(filtered_rides)
dim(filtered_rides)
head(filtered_rides)
str(filtered_rides)
summary(filtered_rides)
```

## Familiarize and Verify Expected Values
View unique column values. This both familiarizes the values as well as verifies that the data is consistent.
```{r verify unique, results='markup'}
unique(filtered_rides$rideable_type)
unique(filtered_rides$member_casual)
```
### Filter Out Bad Datetime Data
Verify the started_at datetime is before the ended_at datetime. If the started_at datetime is equal to or after the ended_at datetime, this will cause problems in analyzing trip durations. We will filter out any rows with started_at datetimes that occur at the same time as or after ended_at datetimes. The filtered rows will be saved to a separate data frame for review.
```{r filter bad datetime, results='markup'}
removed_rides <- filtered_rides %>%
  filter(filtered_rides$started_at >= filtered_rides$ended_at)
filtered_rides <- filtered_rides %>%
  filter(filtered_rides$started_at < filtered_rides$ended_at)
```

### Verify the Data Frame After Filtering and Removing Rows
```{r verify dimensions, results='markup', message=TRUE}
cat('Total Original Rows: ', nrow(all_rides), '\nTotal Filtered Rows: ', nrow(filtered_rides), '\nTotal Removed Rows:  ', nrow(removed_rides), '\nTotal Filtered & Removed Rows: ', nrow(filtered_rides)+nrow(removed_rides))
```
Sample removed data:
```{r results='markup'}
head(removed_rides)
```
Sample filtered data:
```{r results='markup'}
head(filtered_rides)
```
## Save the Data
Once we have addressed all errors and inconsistencies, we can conclude the filtering process. Now we will export the data into their respective CSV files for the Analyze Phase.

### Export the Original Data
```{r export original data}
write.csv(all_rides, file='original_rides.csv', row.names=FALSE)
```

### Export the Cleaned Data
```{r export cleaned data}
write.csv(filtered_rides, file='cleaned_rides.csv', row.names=FALSE)
```

### Export the Bad Data
```{r export bad data}
write.csv(removed_rides, file='removed_rides.csv', row.names=FALSE)
```