---
title: "CGDA - Case Study 1 - Analyze Data"
output: html_document
date: "2023-03-29"
---
# Analyze the Data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results = 'hide')
knitr::opts_chunk$set(error = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

## Load Packages
```{r load packages}
library(tidyverse)
library(lubridate)
library(scales)
```

## Load CSV File
We will load the previously-cleaned ride data.
```{r load csv into data frame}
all_rides <- read_csv("cleaned_rides.csv")
```

## Quick Look
Take a quick look at the structure of the data frame.
```{r quick look, results='markup'}
glimpse(all_rides)
```

## Aggregate Data & Perform Calculations for New Columns
Aggregating/Deaggregating data and performing calculations to create new views of the same data can aid in finding trends.

### Deaggregate Ride Dates
We will split the started_at datetime into separate fields for further analysis.
```{r deaggregate started_at}
all_rides$date <- as.Date(all_rides$started_at)
all_rides$month <- format(as.Date(all_rides$date), "%m")
all_rides$day <- format(as.Date(all_rides$date), "%d")
all_rides$year <- format(as.Date(all_rides$date), "%Y")
```

### Add a Weekday Column
We will add a weekday column calculated from the ride date.
```{r add a ride weekday}
all_rides$day_of_week <- format(as.Date(all_rides$date), "%A")
```

### Add a Ride Duration Column
We will calculate the length of the ride in seconds and add it to a new column, ride_length
```{r add a ride duration}
all_rides$ride_length <- difftime(all_rides$ended_at, all_rides$started_at)
# convert the ride_length from factor to numeric so we can run calculations
all_rides$ride_length <- as.numeric(as.character(all_rides$ride_length))
```

## Verify the Calculated Data
We will verify the new structure and sample data from the data frame.
```{r verify changes, results='markup'}
glimpse(all_rides)
head(all_rides)
```

## Conduct Analysis of the Data

### View Summary Data of ride_length
```{r analysis-summary ride_length, results='markup'}
summary(all_rides$ride_length)
```

### Compare ride_length Between Members and Casual Riders
```{r analysis-compare ride_length, results='markup'}
aggregate(all_rides$ride_length ~ all_rides$member_casual, FUN=mean)
aggregate(all_rides$ride_length ~ all_rides$member_casual, FUN=median)
aggregate(all_rides$ride_length ~ all_rides$member_casual, FUN=max)
aggregate(all_rides$ride_length ~ all_rides$member_casual, FUN=min)
```

### Compare ride_length by Number of Rides & Weekday Between Members and Casual Riders
```{r analysis-compare duration with weekday between rider types, results='markup'}
all_rides %>%
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarize(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(weekday, member_casual)
```

### Compare Total Rides & Total Duration Between Members and Casual Riders
```{r analysis-compare totals between rider types}
all_rides %>%
  group_by(member_casual) %>%
  summarize(number_of_rides = n(), average_duration=sum(ride_length)) %>%
  arrange(member_casual)
```

## Create Plots
We will create some plots to take a quick look at the data for visual trends.

### Plot Number of Rides by Weekday & Rider Type
```{r plot num rides by weekday/riderType, results='markup'}
all_rides %>%
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarize(number_of_rides = n(), average_duration=mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x=weekday, y=number_of_rides, fill=member_casual)) + geom_col(position="dodge") + scale_y_continuous(labels = scales::label_number_si())
```

### Plot Average Duration by Weekday & Rider Type
```{r plot avg duration by weekday/riderType, results='markup'}
all_rides %>%
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarize(number_of_rides = n(), average_duration=mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x=weekday, y=average_duration, fill=member_casual)) + geom_col(position="dodge") + scale_y_time()
```

### Plot Number of Rides by Month & Rider Type
```{r plot num rides by month/riderType, results='markup'}
all_rides %>%
  mutate(monthyear = floor_date(date, unit="month")) %>%
  group_by(member_casual, monthyear) %>%
  summarize(number_of_rides = n(), average_duration=mean(ride_length)) %>%
  arrange(monthyear, member_casual) %>%
  ggplot(aes(x=monthyear, y=number_of_rides, fill=member_casual)) + geom_col(position="dodge") + scale_y_continuous(labels = scales::label_number_si()) + scale_x_date(NULL, date_labels='%b %y', breaks='month')
```

### Plot Average Duration by Month & Rider Type
```{r plot avg duration by month/riderType, results='markup'}
all_rides %>%
  mutate(monthyear = floor_date(date, unit="month")) %>%
  group_by(member_casual, monthyear) %>%
  summarize(number_of_rides = n(), average_duration=mean(ride_length)) %>%
  arrange(monthyear, member_casual) %>%
  ggplot(aes(x=monthyear, y=average_duration, fill=member_casual)) + geom_col(position="dodge") + scale_y_time() + scale_x_date(NULL, date_labels='%b %y', breaks='month')
```

## Export Analysis and Data Frame with Calculated Values
This markdown will be exported for an analysis summary.

Next we will export the data frame with the new calculated values.
```{r export csv, eval=FALSE}
write.csv(all_rides, file='analyzed_rides.csv', row.names=FALSE)
```